---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r echo=TRUE}
library(dplyr)
library(ggplot2)
activity <- read.csv("activity.csv", na.strings="NA")
```

## What is the mean total number of steps taken per day?

We group the activity by day ('date' is a factor here) and calculate the sum of 
steps for each day:

```{r echo=TRUE}
activity.day <- group_by(activity, date)
activity.day.total <- summarise(activity.day, sum(steps))
```

This gives us the following histogram:

```{r echo=TRUE}
names(activity.day.total) <- c('date', 'total.steps')
hist(activity.day.total$total.steps, breaks=10, 
     main="Total number of steps per day", xlab="Steps per day")
```

### Mean and median

```{r echo=TRUE}
steps.mean <- format(mean(activity.day.total$total.steps, na.rm=TRUE), 
                     scientific=F)
steps.median <- format(median(activity.day.total$total.steps, na.rm=TRUE), 
                       scientific=F)
```

The mean of the total number of steps taken per day is `r steps.mean` and the 
median is `r steps.median`.

## What is the average daily activity pattern?

The average daily activity pattern looks as follows:

``` {r echo=TRUE}
activity.timeseries <- group_by(activity, interval)
activity.timeseries.mean <- summarise(activity.timeseries, mean(steps, 
                                                                na.rm=TRUE))
names(activity.timeseries.mean) <- c('interval', 'steps.mean')
ggplot(activity.timeseries.mean, aes(interval, steps.mean)) + 
  geom_line() +
    xlab("5 minute Intervals") +
    ylab("Average Number of Steps")
```

``` {r echo=T}
max.steps <- max(activity.timeseries.mean$steps.mean)
activity.timeseries.mean.sorted <- 
  activity.timeseries.mean[rev(order(activity.timeseries.mean$steps.mean)),]
max.steps.interval <- activity.timeseries.mean.sorted[1,1]
```

On average across all the days in the dataset, the highest number of steps 
(`r max.steps`) is taken in interval __`r max.steps.interval`__.

## Imputing missing values

Calculating the missing values for each column, we see that only the 'steps' 
column is missing data.

``` {r echo=T}
apply(is.na(activity),2,sum)
```

To fill in the missing data, we add the mean number of steps per interval for 
every interval that is missing data (found in the `activity.timeseries.mean` 
dataframe we used in the previous section):

``` {r echo=T}
activity.nomissing <- cbind(activity,activity.timeseries.mean[2])
activity.nomissing$steps[is.na(activity.nomissing$steps)] <- 
  activity.nomissing$steps.mean[is.na(activity.nomissing$steps)]
drop <- c('steps.mean')
activity.nomissing <- activity.nomissing[,!names(activity.nomissing) %in% drop]
apply(is.na(activity.nomissing),2,sum)
```

### Plotting the histogram for filled in data

We now follow the same procedure as in the first section, first grouping the 
activity by day ('date' is again a factor) and calculating the sum of steps for 
each day:

```{r echo=TRUE}
activity.nomissing.day <- group_by(activity.nomissing, date)
activity.nomissing.day.total <- summarise(activity.nomissing.day, sum(steps))
```

This gives us the following histogram:

```{r echo=TRUE}
names(activity.nomissing.day.total) <- c('date', 'total.steps')
hist(activity.nomissing.day.total$total.steps, breaks=10, 
     main="Total number of steps per day", xlab="Steps per day")
```

### Mean and median with filled in data

```{r echo=TRUE}
steps.mean.nomissing <- format(mean(activity.nomissing.day.total$total.steps), 
                               scientific=F)
steps.median.nomissing <- format(median(activity.nomissing.day.total$total.steps), 
                                 scientific=F)
```

The mean of the total number of steps taken per day when data has been filled in 
is `r steps.mean.nomissing` and the median is `r steps.median.nomissing`. Note 
that the mean is exactly the same as before after this operation, but the median 
is now ever so slightly higher (indeed it has become equal to the mean).

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}
activity$date <- strptime(activity$date, format="%Y-%m-%d")
weekdays <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activity$time.of.week <- factor((weekdays(activity$date) %in% weekdays), 
                                levels=c(FALSE, TRUE), 
                                labels=c('weekend', 'weekday'))
activity$date <- as.character(activity$date)
activity.wday <- group_by(activity, interval, time.of.week)
activity.wday.timeseries.mean <- summarise(activity.wday, mean(steps, 
                                                               na.rm=TRUE))
names(activity.wday.timeseries.mean) <- c('interval', 'time.of.week', 
                                          'steps.mean')

ggplot(activity.wday.timeseries.mean, aes(as.integer(interval), steps.mean)) +
    geom_line() +
    xlab("5 minute Intervals") +
    ylab("Average Number of Steps") +    
    facet_grid(time.of.week ~ .)
```